```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```

Red Wine Quality by Yuan Cao
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

library(ggplot2)
library(GGally)

```

```{r echo=FALSE, Load_the_Data}

wine = read.csv("wineQualityReds.csv")

```

I will be exploring techniques of data analysis and prediction using the red
wine quality data used by Cortez et al in their 2009 paper: Modeling wine
preferences by data mining from physicochemical properties. The article appeared
on the November issue of Decision Support System.

There has been a number of publications that questioned the reliability of human
judges when it comes to juding wine quality. It would be interesting to see if 
human judgements has some basis in chemical properties of wines.

# Univariate Exploration and Plots Section

## Overview of the data

```{r echo=FALSE}

names(wine)

```

The physical and chemical properties of the wine under consideration are as
follows:

Fixed Acidity - concentration of tartaric acid (g/L)
Volatile Acidity - concentration of acetic acid (g/L)
Citric Acid - concentration of citric acid (g/L)
Residual sugar - concentration of residual sugar (g/L)
Chlorides - concentration of sodium chloride (g/L)
Free Sulfur Dioxide - concentration of free sulfure dioxide (mg/L)
Total Sulfur Dioxide - concentration of total sulfure dioxide (mg/L)
Sulphates - concentration of potassium sulphate (g/L)
Alcohol - amount of alcohol in the red win (% per volume)
Density - density of the red wine (g/ml)
pH - pH of the red wine

```{r echo=FALSE}

str(wine)

```

X marks the index of the wine data. Quality is the score given by judges, which
can take on an integer value between 0 and 10. Other variables are continuous.

```{r echo=FALSE}

summary(wine)

```

Not counting X, which is the index, most variables takes on values that are
reasonable. However, I am concerned that quality score. While it can take on
values from zero to ten, it only has a range from three to eight.

```{r echo=FALSE}

ggplot(aes(x=quality), data=wine) + geom_bar() +
  scale_x_continuous(breaks=seq(0, 10, 1)) +
  ggtitle("Wine Quality by Count in Wine Data")

```

```{r echo=FALSE}

summary(wine$quality)

```

It seems most wine have a quality score of 5 or 6. Any wine scoring above or
below are uncommon. There are only a few samples of extreme good and extreme
bad wines. It will be difficult to make a convincing arguement regarding what
would make a great or horrible red wine.

While quality score can be considered categorical, it will be difficult to
assess what makes a wine a 5 as opposed to 8 due to the paucity of data in some
categories. Instead, I will treat quality data as if it was continuous. I hope
this would allow me to draw a progression from worse to better red wines and
show how their chemical composition would change as the result. Since I have
few data point at either extreme, I hope that such a progression would lend
more credibility to my conclusions.



## Macro Variables

Before examine the chemicals under consideration. It might be good to get an
overall sense of what we are trying to analyze by looking at the macroscopic
characteristics of the data.

### Density

The most common substance in a red wine is water. Under standard conditions, the
concentration of water is 1 g/ml. Water is considered an especially dense liquid
because of water molecule's (H2O's) ability to form hydrogen bonds with other
water molecules. As additional substances are added to water, it generally
weakens water's potential to form Hydrogen bonds. 

```{r echo=FALSE}

ggplot(aes(x=density), data=wine) + geom_histogram() +
  xlab("Density (g/ml)") +
  ggtitle("Density Distribution of Red Wine")

```

As expected, red wine generally seems to be slightly less dense than water. The
density of red wine seems to be normally distributed.

```{r}

summary(wine$density)

```

### pH

Most of the substances under consideration are acids of some sort. I would
expect red wines have a lower pH than water. 

```{r echo=FALSE}

ggplot(aes(x=pH), data=wine) + geom_histogram() +
  ggtitle("pH Distribution of Red Wine")
```

The pH value of red wine seem to form a normal distribution.

```{r echo=FALSE}

summary(wine$pH)

```


### Alcohol

It is perhaps obvious that we drink wine for the alcohol contend of the drink.
It would be interesting to see if alcohol content for red wine is also normally
distributed.

```{r echo=FALSE}

ggplot(aes(x=alcohol), data=wine) + geom_histogram() +
  xlab("Alcohol (% by Volume)") + 
  ggtitle("Alcohol Distribution of Red Wine")
```

For the first time, we see a skewed graphed. This might be an indication that
alcohol plays a role in the quality accessment of the wine.

# Univariate Analysis

In the preliminary exploration, I concentrated on the big picture view. The
various types of acids in the wine all contribute to lower the pH of the wine.
Depending on the chemicals involved, they will affect the density of the wine.
As expected, they mostly have normal distributions.

I am most concerned about the relatively few examples of really good or really
poor wine. Although the scale of possible wine quality scores allowed for any
value between 1 and 10. The highest score was 8, and the lowest score was 3.
There were less than 30 data points (out of 1599 observations) in the extreme
ends of the quality values. Most data are clumped in the middle. It will be
hard to justify any prediction regarding characteristics of very good (9s, 10s)
or very bad (1s, 2s) wine when there are no samples of those types of wine in
the data set.

In the subsequent exploration and analysis, I hope to see if there might be a
progression of change in chemical properties of red wine that would explain the
change in quality accessment. If the trend is strong enough, I hope it would
lend me credence in drawing conclusions for what makes a great or poor red wine.

# Bivariate Exploration and Plots Section

## Macroscopic Variables and Quality Assessment

Based on univariate exploration, I would expect that the macroscopic variables
that we explore would contribute little to the quality score, unless there is a
strong correlation between them and the quality score. 

### Density

```{r echo=FALSE}

ggplot(aes(x=density), data=wine) + geom_histogram() + xlab("Density (g/ml)") +
  facet_grid(~quality) +
  ggtitle("Density Distribution of Red Wine Split by Quality")

```

Density seem to be normally distributed. The nice bell shaped curve seemed to
break down at the higher and lower scores due to lack of samples.

### pH

```{r echo=FALSE}

ggplot(aes(x=pH), data=wine) + geom_histogram() +
  facet_grid(~quality) +
  ggtitle("pH Distribution of Red Wine Split by Quality")
```

Again, pH seems to be normally distributed for wines in each quality group.

### Alcohol

```{r echo=FALSE}

ggplot(aes(x=alcohol), data=wine) + geom_histogram() +
  xlab("Alcohol (% by Volume)") +
  facet_grid(~quality) +
  ggtitle("Alcohol Distribution of Red Wine split by Quality")
```

If we look at the alcohol content by quality categories, it's hard to see an
obvious pattern. However, it might make sense after further analysis.

### Macroscopic Variables and Quality Assessment Overview

```{r echo=FALSE}

plotter_macro <- subset(wine, select=c(density, pH, alcohol, quality))
ggpairs(plotter_macro)

```

As expected, most variables are only very weakly correlated with quality scores.
However, as we saw from univariate exploration, alcoholic content does seem to
affect the quality score more than the other variables. It would be worthwhile
to look into this further in the coming sections.

## Acids

The data looked at a few different acids.

```{r echo=FALSE}

plotter_acids <- 
  subset(wine, 
         select=c(fixed.acidity,
                  volatile.acidity,
                  citric.acid,
                  quality))
ggpairs(plotter_acids)

```

While it would be good to keep citric acid and volatile acidity in mind, it
seems the various acids only weekly correlate with quality.

## Sulfur

The next category of substances are related to sulfur. It's measured by free
sulfur dioxide, total sulfur dioxide and sulphates. 

```{r echo=FALSE}

plotter_sulfur <- 
  subset(wine, 
         select=c(free.sulfur.dioxide, total.sulfur.dioxide, 
                  sulphates, quality))
ggpairs(plotter_sulfur)

```

However, it still seems like there are only weak correlation between sulfur
measurements and quality score. At this point, it seems likely that quality
score is related to a combination of many different factors.

## Remaining Variables

Only residual sugar and chlorides remain unexplored. Since residual sugar is
related to the state of the red wine at the end of the fermentation process, I
think it would be interesting to look at it along with alcohol content. I expect
a negative correlation between the two variables.


```{r echo=FALSE}

plotter_other <- 
  subset(wine, select=c(chlorides, residual.sugar, alcohol, quality))
ggpairs(plotter_other)

```

Although my suspicious that red wine quality is most likely due to a
combinatorial effect of many different variables is confirmed, it's still a
shock that it seems residual sugar level is not at all related to alcohol level.

## Best of Bred

It would be interesting to see if alcohol content is related volatile acidity,
since they seems to be the two factors that has the greatest individual impact
on a red wine's quality score.

```{r echo=FALSE}

plotter_best <- 
  subset(wine, select=c(volatile.acidity, alcohol, quality))
ggpairs(plotter_best)

```

It is interesting that there is only a small negative correlation between the
two variables. I would have expected them to have a much higher negative
correlation. This seems to further indicate that quality score is based on a
more complex interaction between the various factors, or there is no underlying
physical or chemical foundation for judges' assessment of red wines.

# Bivariate Analysis

```{r echo=FALSE}

ggplot(aes(x=alcohol, y=volatile.acidity, color=factor(quality)), data=wine) +
  geom_point() + scale_color_brewer(palette="RdYlGn") + theme_dark() +
  xlab("Alcohol (% by Volume)") + ylab("Volatile Acidity (acetic acid, g/L)") +
  ggtitle("Relationship between Alcohol Level and Volatile Acidity")

```

Two factors stands out among the different types of data collected. Alcohol
content seems to have the biggest positive correlation with the quality score
of the red wine in the study. Volatile acidity seems to have the biggest
negative correlation with the quality score. The other factors seems to play
marginal roles.

It seems reasonable that people would prefer to have a stronger drink - after
all, if it's not alcohol we are after when we drink wine, we might as well drink
grape juice. It does seem surprising that there seems to be one major mistakes
one needs to avoid to make good red wine - having too much volatile acidity.

# Multivariate Exploration and Plots Section

I would like to explore the relationship between the various acids and the
quality score of a red wine.

## Volatile Acidity

Since having a high volatile acidity seems to be the main factor that would
drag down a wine's quality score, it would be interesting to see how it relates
with other factors. In particular, I am interested to explore its connection
with other acids and the pH level of the wine.

```{r echo=FALSE}

wine$minus_pH <- wine$pH * -1

plotter_acids_ph <- 
  subset(wine,
         select=c(fixed.acidity, volatile.acidity,
                  citric.acid, minus_pH, quality))
ggpairs(plotter_acids_ph)

```

To make the analysis simpler, I created the minus_pH variable. It's simply pH
value multiplied by negative one. This transformation results in higher
minus_pH value (closer to zero) being more acidic. A higher concentration of
acids results in higher minus_pH value.

It seems strange that while fixed acidity and citric acid concentration both
positively correlated with minus_pH, there is a slight negative correlation
between minus_pH and volatile acidity.

It seems to imply that at higher acidity levels, there tends to be less volatile
acidity. Since volatile acidity is negatively correlated with quality scores, it
would imply that more acidic wine should be favored. However, minus_pH (pH) does
not seem to be correlated with a red wine's quality score.

This seems to point to the possibility that it's the ratio of acids that
determine the wine quality.

```{r echo=FALSE}

wine$total.acid <- wine$fixed.acidity + wine$volatile.acidity + wine$citric.acid
wine$fixed.acidity.ratio <- wine$fixed.acidity / wine$total.acid
wine$volatile.acidity.ratio <- wine$volatile.acidity / wine$total.acid
wine$citric.acid.ratio <- wine$citric.acid / wine$total.acid

plotter_acid_ratio <- subset(wine, 
                             select=c(fixed.acidity.ratio,
                                      volatile.acidity.ratio,
                                      citric.acid.ratio,
                                      quality))

ggpairs(plotter_acid_ratio)

```

```{r echo=FALSE}

wine$fixed.volatile.ratio <- wine$fixed.acidity / wine$volatile.acidity
wine$citric.fixed.ratio <- wine$citric.acid / wine$fixed.acidity
wine$citric.volatile.ratio <- wine$citric.acid / wine$volatile.acidity
wine$nonvol.volatile.ratio <- 
  (wine$citric.acid + wine$fixed.acidity) / wine$volatile.acidity

plotter_acid_pair <- subset(wine, select=c(fixed.volatile.ratio, 
                                           citric.fixed.ratio,
                                           citric.volatile.ratio,
                                           nonvol.volatile.ratio,
                                           quality))

ggpairs(plotter_acid_pair)

```

The best feature that I could come up with seems to the the ratio of
non-volatile acidity (fixed + citric acids) to volatile acidity. Interestingly,
it has a correlation with fixed to volatile acidity ratio of 1. Although fixed
to non-volatile acidity ratio has a slightly lower correlation with quality
score, it's small enough that it does not out weigh the benefit of having a
simpler model.

```{r echo=FALSE}

ggplot(aes(x=volatile.acidity, y=quality, color=fixed.volatile.ratio),
       data=wine) + geom_point() + geom_jitter() +
  xlab("Volatile Acidity (acetic acid, g/l)") + ylab("Quality") + 
  ggtitle("Beyond Volatile Acidity (Fixed to Volatile Acidity Ratio)") +
  scale_color_continuous(low="brown", high="yellow",
                         name="Fix-Volatile Acid Ratio")

```

Citric acid concentration could perhaps be better used as another layer on a
potential model for assessing whether human judge's scoring might have a
underlying basis in chemistry. However, it does not seem as important as the
fixed to volatile acidity ratio.

```{r echo=FALSE}

ggplot(aes(x=volatile.acidity, y=quality, color=citric.acid.ratio), data=wine) +
  geom_point() +
  geom_jitter() +
  xlab("Volatile Acidity (acetic acid, g/l)") + ylab("Quality") +   
  ggtitle("Beyond Volatile Acidity (Citric Acid Level to Volatile Acidity Ratio)") +
  scale_color_continuous(low="brown", high="yellow", name="Citric Acid Ratio")

```


# Multivariate Analysis

I found two factors that most affect the quality scoring of a red wine: alcohol
level and volatile acidity. I felt it was clear why alcohol level would play a
role in determining a red wine's quality score. So, I focused my analysis on the
nature of the acidity found in red wines and how it impact quality scoring of
red wines.

Beyond a preference against volatile acidity, judges seem to favor a higher
ratio of non-volatile to volatile acidity ratio. Citric acid concentration did
not seem to have as large an affect as I originally thought. I believe if I were
to build a model to predict quality score based on the chemical and physical
properties of a red wine, fixed to volatile acidity ratio would serve as a
better variable.

### Modeling of Red Wine

```{r}

library(lattice)
library(MASS)
library(memisc)

m1 <- lm(quality ~ alcohol, data=wine)
m2 <- update(m1, ~ . + volatile.acidity)
mtable(m1, m2)

```

While grading red wines are complicated, it's interesting to see what a not
insignificant part of the process can be attributed to two variables. By
including more variables, it's possible to increase the predictive power of the
model. However, I believe the model will be less robust. There are only a few
samples of really good and really bad wines. A complex model is more likely to
overfit on those few samples, and obscure the underlying causes for what makes
a red wine good.

The main focus of this series of analysis is to assess if judges' quality score
assessment has a basis in the underlying physical or chemical properties
of a red wine. I believe this indicates that there is such a basis, but it is
not a major contributing factor to red wine quality scores.

------

# Final Plots and Summary

After our analysis, it might be interesting to shift focus toward something more
lighthearted. At the end of the day, all we want is to be able to pick good red
wines for ourselves. Although red wine scoring is a complicated process, there
are a couple of major themes that would allow someone to pick up a reasonable
bottle with relatively few pieces of information.

### Plot One
```{r echo=FALSE, Plot_One}
ggpairs(plotter_best)
```

### Description One

There are two most important single factors that seem to affect human judges.
One is the alcohol level in a red wine. The other is the volatile acidity in a
red wine.

In practical terms, if you know nothing about two bottles of red wine, you could
do worse than pick the one with a higher alcoholic content. If you could only
know one more piece of information about the red wine, find out its volatile
acidity.

### Plot Two
```{r echo=FALSE, Plot_Two}

ggplot(aes(x=alcohol, y=volatile.acidity, color=factor(quality)), data=wine) +
  geom_point() + scale_color_brewer(palette="RdYlGn") + theme_dark() +
  xlab("Alcohol (% by Volume)") + ylab("Volatile Acidity (acetic acid, g/L)") +
  ggtitle("Relationship between Alcohol Level and Volatile Acidity")

```

### Description Two

Luckily, it seems like red wine that has higher alcoholic content seems to have
a lower volatile acidity. This is a weak effect, but simply looking at the
alcohol content can tell you quite a lot.

### Plot Three
```{r echo=FALSE, Plot_Three}

ggplot(aes(x=volatile.acidity, y=quality, color=fixed.volatile.ratio),
       data=wine) + geom_point() + geom_jitter() +
  xlab("Volatile Acidity (acetic acid, g/l)") + ylab("Quality") + 
  ggtitle("Beyond Volatile Acidity (Fixed to Volatile Acidity Ratio)") +
  scale_color_continuous(low="brown", high="yellow",
                         name="Fix-Volatile Acid Ratio")

```

### Description Three

If you have the complete data of a set of red wines, but not enough time to do
an indepth analysis, look at the fixed to volatile acid ratio, after you have
find out the wine's alcohol level and volatile acidity.

------

# Reflection

None of the factors seems be a silver bullet for predicting red wine quality
scores. It might be due to individual natures of people's tastes. As other
publications have already pointed it out, it might also be due to the
inconsistency of the judges, which too, is a human failing. In addition, there
are few samples that has a high or low quality score. This limits how robust any
model can be.

I believe there are two factors that stands out: alcohol level and volatile
acidity. They seems to be the factors that has the greatest effect on the
quality score of a red wine.

With the paucity of data points on each end of quality score, I think another
interesting way to analyze the data might be to eliminate the middle data points
and only look at the extreme data points. It might be interesting to see if
that might make some trends in the data more obvious. Then, we can try to see
if the trend holds when we test it using all of the data.